---
title: "p8105_hw2_ly2633"
author: "Leila Yan"
date: "2024-09-25"
output: github_document
---

```{r setup}
library(tidyverse)
```
## Problem 1
Read in the dataset
```{r}
transit_df = 
  read_csv("Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>%
  mutate(entry = case_match(entry, 
                            "YES" ~ TRUE,
                            "NO" ~ FALSE))
```
Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy? How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here. How many stations are ADA compliant? What proportion of station entrances / exits without vending allow entrance?

Answers:



Reformat data so that route number and route name are distinct variables. 
```{r}
transit_df_long=
  transit_df %>%
  mutate(across(starts_with("route"), as.character)) %>%
  pivot_longer(
    route1:route11, 
    names_to = "route_number", 
    values_to = "route_name", 
    values_drop_na = TRUE
    )
```

How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r}
stations_A_train =
  transit_df_long %>%
  filter(route_name == "A")


distinct_stations_A_train =
  stations_A_train %>%
  distinct(station_name) %>%
  nrow()

ada_compliant_stations_A_train =
  stations_A_train %>%
  filter(ada == TRUE) %>%   # Filter for ADA compliant stations
  distinct(station_name) %>%  # Get distinct station names
  nrow()
```

Answers:


## Problem 2

## Problem 3
