---
title: "p8105_hw2_ly2633"
author: "Leila Yan"
date: "2024-09-25"
output: github_document
---

```{r setup}
library(tidyverse)
```
## Problem 1
Read in the dataset
```{r}
transit_df = 
  read_csv("Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>%
  mutate(entry = case_match(entry, 
                            "YES" ~ TRUE,
                            "NO" ~ FALSE))
```
Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy? How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here. How many stations are ADA compliant? What proportion of station entrances / exits without vending allow entrance?

Answers:



Reformat data so that route number and route name are distinct variables. 
```{r}
transit_df_long=
  transit_df %>%
  mutate(across(starts_with("route"), as.character)) %>%
  pivot_longer(
    route1:route11, 
    names_to = "route_number", 
    values_to = "route_name", 
    values_drop_na = TRUE
    )
```

How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r}
stations_A_train =
  transit_df_long %>%
  filter(route_name == "A")


distinct_stations_A_train =
  stations_A_train %>%
  distinct(station_name) %>%
  nrow()

ada_compliant_stations_A_train =
  stations_A_train %>%
  filter(ada == TRUE) %>%   # Filter for ADA compliant stations
  distinct(station_name) %>%  # Get distinct station names
  nrow()
```

Answers:
16 stations that serve the A train are ADA compliant.



## Problem 2

Read and clean the Mr. Trash Wheel sheet:
- specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
- use reasonable variable names
- omit rows that do not include dumpster-specific data
- round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
library(readxl)
mr_trash_wheel_df <- read_excel("Data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel") %>%
  janitor::clean_names() %>%
  filter(!is.na(`Dumpster`)) %>%
  select(dumpster, date, Weight_tons,`Plastic Bottles`, `Polystyrene`, `Cigarette Butts`, 
         `Glass Bottles`, `Plastic Bags`, `Wrappers`, `Sports Balls`)




```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}
prof_trash_wheel_df <- read_excel("Data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel")
gwynnda_trash_wheel_df <- read_excel("Data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel")
```



## Problem 3

```{r}

```

